Debugging by {\huge\bf\textit{lastChange}}

### 2011-03-10 trunk/papers/fse2011/sig-alternate.tex

\begin{figure}[htp]
\centering 
\includegraphics [width=.48\textwidth] {6-foo-changes1.png} %[width=.48\textwidth][scale=0.75]
\caption{A hypothetical list of  change events for a property \texttt{foo}. 
Each change event adds a column with the id of the object changed, the call stack,
and some program state such as local variable values. 
At the reproduction point  we determine which id  corresponds to 
object \texttt{bar} and read out column 4,  the last
  change of \texttt{bar.foo}. Column 3 is also a change of
  the object we want to study, id 1010, but it is not the last change;
  Column 5 is also a change of a property \texttt{foo} but it is not for
  the object we are interested in.}
\label{fig:foo-changes1}
\end{figure}

\subsection{{\large\bf\textit{lastChange}} on variable} 
In JavaScript, every frame has a scope chain and every available
variable in the frame comes from one of the scopes in the frame's
scope chain. Once the developer asks for the last change of a variable with name 
\texttt{foo} at the reproduction point, the debugger first determines the
variable's scope as follows: it iterates over the scopes in the scope
chain and the first scope which has a variable with the same name is
the variable's scope. There are five different scope types: local,
global, \texttt{with}, closure and \texttt{catch}. We explain these
cases in two groups.

@@

\subsubsection{local, closure and \texttt{catch} scopes}
Local scope refers to the most nested scope in the scope
chain which contains the local variables. Closure scope refers to the 
scope which is created for a nested function and contains variables 
defined in the outer block. Catch scope is the scope created in the catch
block of try-catch statements and contains the exception
variable. These scopes are not necessarily regular JavaScript
objects. Scope object is usually a transient copy of a native object. 
Therefore, to track changes to a variable in these scopes we
employ a different approach.

\begin{figure}[htp]
%\begin{verbatim}
\lstset{basicstyle=\scriptsize}
\begin{lstlisting}[language=myLang, framerule=0pt]
  function f(){
    var x = 0;
    x++;
    if (!stop) f();
  }

Sample Trace:

   f()
A  |  x changes, scope 1 
   |  f()
   |  |  x changes, scope 2
   |  |  f()
B  |  |  |  x changes, scope 3 
C  |  ... , scope 1
\end{lstlisting}
%\end{verbatim}
\caption{A recursive call trace illustrating the scope
  id. The lines in the bottom half of the diagram simulate a trace of the change events for the variable \texttt{x} as the function \texttt{f()} calls itself. Each call creates a scope; eventually when the variable \texttt{stop} is changed by an external process we return from the recursion. The lines marked A, B, and C are discussed in the text.}
\label{fig:recursive}
\end{figure}

Having the scope chain and the source code, we can map every scope to % perhaps needs more explanation, how?
a code block, enclosed the executing code. In
JavaScript, a code block can be identified by the file url and  
the block's first instruction program counter. Given this information, the debugger is
able to recognize the code block in loaded scripts or once it is loaded.
Similar to \textit{lastChange} on object property, we define two basic operations: 
\texttt{scopeId()} and \texttt{setVariableChangeHook()}. The first one, 
given a scope returns an integer as the scope's identifier. The second one, 
given a callback function, a code block and a variable name-which is defined inside the block, the function is called
whenever the variable is changed.

@@

\begin{figure}[htp]
\centering 
\includegraphics[width=.40\textwidth]{7-foo-changes2.png}
\caption{A hypothetical list of variable \texttt{foo} change events. Each column of the list indicates a change event; for each
change event the scope id returns by \texttt{scopeId()} is recorded along with call stack and program state information. 
   The last column having the scope id \texttt{foo} at the reproduction point indicates 
  the last change.}
\label{fig:foo-changes2}
\end{figure}

\subsection{{\large\bf\textit{lastChange}} on {\large\bf\textit{lastChange}}}
The \textit{lastChange} algorithm records changes by id (either object or scope id), then reads out the last change when we
arrive at the reproduction point and discover the id of the requested value.
When we perform \textit{lastChange} based on a previous \textit{lastChange} the query algorithm 
must retain additional information. Consider the following example:

@@

%---------------------------------------------------------------------------------------------------
\section{JavaScript implementation}
To verify the \textit{lastChange} algorithm we
implemented\footnote[1]{http://code.google.com/p/querypoint-debugging} it in an extension to the Firebug
JavaScript debugger\footnote[2]{http://getfirebug.com}. %\cite{Firebug}. %\cite{firebug-version-1.6}. 
Firebug itself is an extension of the Firefox browser. %\cite{Firefox}. %\cite{firefox-version-3.6}
The Firefox JavaScript engine provides a JavaScript debugging interface \cite{JSD} and 
\textit{Querypoint} is developed over this interface. Our prototype implementation defines our four primitive operations in JavaScript code and using techniques which are cumbersome and comparatively slow to execute. However the JavaScript prototype is
easy to explore, change, and share with others for feedback.
  A professionally useful debugger would implement these primitive operations within the JavaScript engine.

\begin{figure}[htp]
\centering 
\includegraphics[width=.48\textwidth]{8-lastchange-lastchange.png}
\caption{The list of change events stored for locating point B, the
  \textit{lastChange} of \texttt{bar.x} at the reproduction point, and
  point C, the \textit{lastChange} of \texttt{baz.y} at point B.}
\label{fig:lastchange-lastchange}
\end{figure}

\subsection{{\normalsize\bf\texttt{objectId()}} operation}
\texttt{objectId(obj)} first checks the argument \texttt{obj} for a property \texttt{\_objectId}.
It returns the value if this property is already defined,
otherwise it generates a new id and sets this property. The value of the id is simply an integer incremented 
for each new \texttt{\_objectId} needed. This use of a regular object property  might change
the program behaviour. For example it will apear as an extra property in 
\texttt{for(property in object)} loops. 
The next version of Firefox\footnote[3]{https://developer.mozilla.org/en/JavaScript/Reference\\/Global\_Objects/Object/defineProperty} provides an alternative via \texttt{defineProperty} a standard JavaScript
function which receives 
an argument specifying whether the property is enumerable or not. By setting the enumerable
field \texttt{false} for \texttt{\_objectId}, this property will not
apear in \texttt{for(property in object)} loops and therefore has no
effect on the program execution. Note that \texttt{\_objectId} is
not set for all objects but only those objects that need an id.

@@

For our purposes, the \texttt{watch()} function only covers the case
of global object properties. At the beginning of execution, no object
excepting the global object and its predefined properties is available.  For 
our \textit{lastChange} prototype we created a version of
\texttt{setPropertyChangeHook()}.  The basic strategy is to get a reference to the object just
after its creation, then use \texttt{watch()} function to monitor
property changes in the object. Setting a flag\footnote[4]{DISABLE\_OBJECT\_TRACE defined in jsdIDebuggerService.idl} into the Firefox 
JavaScript engine, we can get file URL and line number for each object
creation (e.g., myFile.js, line 24). We set a hook on this line
and parse the source code to determine which object was created.

@@

\subsection{{\normalsize\bf\texttt{scopeId()}} operation}
The prototype sets a hook at the beginning of all code blocks needing
a scope id. The scope id is kept as a variable with name 
\texttt{\_scopeId} in the scope. Whenever the hook is hit,  meaning a new scope
is created, \texttt{\_scopeId} is set by calling JavaScript's dynamic compilation 
function \texttt{eval()}. For example, executing \texttt{eval("var \_scopeId = 10")} creates a
variable with name \texttt{\_scopeId} and value \texttt{10} in the
scope of the \texttt{eval()} call, which is our interesting scope. \texttt{scopeId()}
operation returns the value of \texttt{\_scopeId} in the scope.

\subsection{{\normalsize\bf\texttt{setVariableChangeHook()}} operation}
Variables defined in local, closure, and catch scopes are only changed in
their scope; that includes the defining scope and scopes nested whithin them.
For example in Figure~\ref{fig:js-closure}, variable \texttt{myVar} in line 11 can only be
changed in lines 10 to 23. Therefore, after locating the function in
which the variable is defined, it is enough to parse the code inside
the function block and set a hook on all lines where the variable is
assigned a new value. 
We also set a hook at the first line of the function which is
corresponding to the line where the variable is defined. In Figure
~\ref{fig:js-closure}, if the execution is paused at line 17 and the
last change of \texttt{myVar} is queried, two hooks on lines 16 and 17
will be set, but if the execution is paused at line 20 , four hooks on
lines 11, 12, 14, 20 will be set. These two cases are different due
the fact that \texttt{myVar} in \texttt{firstChild} is a local variable
but in \texttt{secondChild} is a closure variable.

@@

There are cases where event indexes fail, but a developer can use knowledge gained from \textit{lastChange}
to formulate a conditional breakpoint.
Consider the example in Figure~\ref{fig:counter-example}. It shows a function which processes an
array. The array contains numbers except one item which is
\texttt{undefined} and it causes a bug in line 30. There is a call to
\texttt{randomPermutation} function in line 11 which randomly
permutes the array item. So at every execution the
\texttt{undefined} item will be in a new place. Calling
\textit{lastChange} on \texttt{x} at the place the bug happens, gives a
point which shows line 14. Although this point exists at every
re-execution but it can not be identified by an event index. A conditional breakpoint on 
the value of the item would succeed and the developer can reason from the 
\textit{lastChange} results to create the condition.

\begin{figure}[htp]
%\begin{verbatim}
\lstset{basicstyle=\scriptsize}
\begin{lstlisting}[language=myLang, framerule=0pt]
10 function randomProcess(array){
11   randomPermutation(array);
12   var x, y;
13   for (var i=0 ; i<array.length ; i++){
14      x = array[i]
...
30      y = x+1;
31   }
32 }
\end{lstlisting}
%\end{verbatim}
\caption{A counter-example for transforming \textit{lastChange} to a conditional breakpoint.}
\label{fig:counter-example}
\end{figure}

\section{User Study}
We supplied four experienced Javascript developers with our prototype lastChange in an 
extended Firebug debugger\footnote[5]{http://ltiwww.epfl.ch/\texttildelow mirghase/lastchange-userstudy}. Following a tutorial and a practice case, we observed as they 
applied both conventional breakpoint and \textit{lastChange} on two small programs we 
provided (Our prototype did not have a user interface to support Sec.~\ref{sec:pausing}).
All four developers successfully applied \textit{lastChange} to the test programs 
and understood how it could help debugging.

We designed the example programs to require two steps of \textit{lastChange}; 
all four users took more steps (9, 13, 12, and more than 30) to find the defect location with breakpoints.  
To find locations for breakpoints two users scrolled through the source, 
another searched the text, a third set a lot of breakpoints to get an understand of the control flow. 
Based on our own experience we expect these strategies represent the kinds of approaches 
developers have available. These operations are time consuming and tedious: all four users 
found the defect location with just two \textit{lastChange} operations. Based on our measurements,
this makes \textit{lastChange} about three times faster than breakpoints for the navigation from 
the reproduction point to the defect
(We recognize that these programs were designed to highlight lastChange and many kinds of debugging 
issues have been hidden by the design of our tests). Our results show that, when a defect 
relates to incorrect values and a developer recognizes this, then the operational 
mechanics of lastChange lead to the defect much more quickly than breakpoints.

@@

The key ingredients in our argument are: 
\begin{enumerate}
   \item developers need an operation like \textit{lastChange}, 
   \item developers can learn to use \textit{lastChange}, 
   \item practical implementations are feasible with modest
invested development time,
   \item in most cases \textit{lastChange} will be much faster than current alternatives,
   \item the worst cases are not more common or more painful than
alternatives.
	 \item the \textit{lastChange} approach can be generalized to other queries.
\end{enumerate}

@@

We have shown how \textit{lastChange} identifies the prior point in
program execution where a program state value changes. Is it something
the developer needs to do?  Since ultimately programs are just
transformation of state values, debugging is ultimately backtracking
to find defects in program state change\cite{Weiser,Zeller}.  When a developer halts a
program on a breakpoint they have three kinds of information: a model
of the program in their mind, a call stack showing what part of the
program is halted, and the state at that point (a combination of the
debugger's view of the program internal state and the output of the
program up to this point of execution). If the model does not match
the call stack or if no value in the viewable program state aligns
with the developers model for this point in execution, the developer
will seek another view by changing the breakpoint or the input
data. If a value appears incorrect, they may have a flash of insight
and know the defect. Otherwise they need to figure out what operation
causes the incorrect value. Here \textit{lastChange} takes over and addresses a key
part of the debugging process.

@@

If \texttt{lastChange} could be useful, can developers figure out how
to use it? While our user study was small, our prototype demonstrates that
\textit{lastChange} can be easily activated by operations on the
graphical representation of erroneous data in a debugger and the 
results can be interpreted by users. We designed the
user interface for the results from \textit{lastChange} to resemble
the results from a breakpoint, with the source code of the change
point highlighted and the state of the program presented the way state
is presented from a breakpoint. Based on this user interface, we
believe developers can start to use \textit{lastChange} with minimal
training. Our users may not be representative, but the key point here
is that the learning curve is small compared with the challenge of learning 
breakpoint debugging in the first place.

While \textit{lastChange} is not a breakpoint, 
all of the assumptions developers already have for breakpoints hold
for \textit{lastChange}. In particular the re-execution is just the
same operation developers use to debug with breakpoints. 
In our own experience, \textit{lastChange} dramatically reduces the work in
setting and removing breakpoints, so work to cause re-execution  become noticable.
Therefore automatic mechanisms
for re-execution will be corresponding more valuable for debugging
with \textit{lastChange} (Our \textit{Querypoint} prototype  
implements a simple form of automatic 
re-execution for both breakpoint and \textit{lastChange}).

@@

We have described our prototype all-JavaScript implementation in Sec. 4. It is adequate for 
exploring the ideas and may even be usable in production. However significant 
improvments can be made. A fully usable
implementation would require access to the object id at the point of
object creation and \texttt{setPropertyChan geHook()}. Many
object-oriented runtimes provide object identifiers and 
provide access to object creation directly or by bytecode instrumentation\cite{JPDA}. 
In the particular case of the Firefox Web browser we used for our prototype, the primary barrier
to practical implementation would likely be integration with the
increasingly sophisticated just in time compilers. In the case of
tracking variable changes, \texttt{setVariableChangeHook()} can be implemented
similar to \texttt{setPropertyChangeHook()} if JavaScript scopes-like regular 
JavaScript objects-support \texttt{watch()} function.

@@

When developers try \textit{lastChange}, will they get results fast
enough to benefit? While we only used our prototype on toy programs,
\textit{lastChange} seemed as fast as breakpoint re-execution.
But what can we say about realistic programs?

@@

We implemented a
simple and effective mechanism allowing developers to control 
overhead: we stored data to a depth of two property accesses 
(e.g. foo.bar.x), then placed a \textit{re-run} button 
in the user interface at the third level. If developers reached the 
third level they can see more data at the cost of one more replay.

For comparison we should use the practical alternative: developers
setting breakpoints. For the vast majority of programs, a developer
will take much more time to set one breakpoint than
\textit{lastChange} would add. But typically the developer may not
guess the point of last change. They must then ponder another
breakpoint and re-execute. Unlike the re-execution we use to 
optimize performance in \textit{lastChange}, 
developers still will not know if the breakpoint they hit is the point of last change
 or even if it is related to the change at all.

@@

Finally we consider the worst cases: what about code that changes
objects in long running loops? Every time through the loop we incur
the call back overhead; if the loop itself has relatively little code
the overhead could be very large; if the loop computation is a
significant fraction of the full program, the slowdown would be
enormous.

These are difficult issues for other techniques as well:  breakpoints are not feasible in these cases and logging
becomes unwieldy. Since automatic logging solutions are highly tuned,
their overhead in this case will likely be much less than
\textit{lastChange}. On the other hand, \textit{lastChange} integrates
with an interactive debugger and detecting that we are in a high
overhead loop is simply a matter of checking our internal counter. In
such unusual cases we may simply offer the developer the option of
studying the loop code then omitting it from \textit{lastChange}
operations. Developers face this issue with any debugger today: occasionally a
debugger causes too much overhead to be useful for debugging. 
The users of \textit{lastChange} may encounter these kinds of cases, but 
that is the nature of debugging: you need different tools for 
different kinds of problems.

\subsection{Generalizations}
Our \textit{lastChange} algorithm can be viewed as a particular interface to a general facility. 
The general facility replays execution, querys the runtime at points of interest during execution,
and analyzes the result at the reproduction point. We have selected one kind of query and analysis
that can be easily integrated in existing debuggers and explained to developers, providing automation of 
the problem of finding the point of last change. We believe other
kinds of queries and analysis can be invented and integrated to automate other aspects of debugging.
Moreover, the query history analysis used to test for non-determinism here may also be the basis for
more effective user interfaces to debugging sessions. The series of cascaded \textit{lastChange} queries becomes 
logical path followed by the developer seeking the defect and this path may be useful if presented in effective user interfaces.

>>>>

\begin{figure}[htp]
\centering 
\includegraphics [width=.48\textwidth] {6-foo-changes1.pdf} %[width=.48\textwidth][scale=0.75]
\caption{A hypothetical list of  change events for a property \texttt{foo}. 
Each change event adds a column with the id of the object changed, the call stack,
and some program state such as local variable values. 
At the reproduction point  we determine which id  corresponds to 
object \texttt{bar} and read out column 4,  the last
  change of \texttt{bar.foo}. Column 3 is also a change of
  the object we want to study, id 1010, but it is not the last change;
  Column 5 is also a change of a property \texttt{foo} but it is not for
  the object we are interested in.}
\label{fig:foo-changes1}
\end{figure}

\subsection{{\large\bf\textit{lastChange}} on variable} 
In JavaScript, every frame has a scope chain and every available
variable in the frame comes from one of the scopes in the frame's
scope chain. Once the developer asks for the last change of a variable with name 
\texttt{foo} at the reproduction point, the debugger first determines the
variable's scope as follows: it iterates over the scopes in the scope
chain and the first scope which has a variable with the same name is
the variable's scope. There are five different scope types: global, local,
closure, \texttt{with} and \texttt{catch}. We explain these
cases in two groups.

@@

\subsubsection{local, closure and \texttt{catch} scopes}
Local scope refers to the most nested scope in the scope
chain which contains the local variables. Closure scope refers to the 
scope which is created for a nested function and contains variables 
defined in the outer block. Catch scope is the scope created in the catch
block of try-catch statements and contains the exception
variable. These scopes are not necessarily regular JavaScript
objects.  
Therefore, to track changes to a variable in these scopes we
employ a different approach.

\begin{figure}[htp]
%\begin{verbatim}
\includegraphics [width=.48\textwidth] {9-scopes.pdf} %[width=.48\textwidth][scale=0.75]

%\lstset{basicstyle=\scriptsize}
%\begin{lstlisting}[language=myLang, framerule=0pt]
%  function f(){
%    var x = 0;	
%    x++;
%    ...
%    if (!stop) f();
%  }

%Sample Trace:

%   f()
%A  |  x changes, scope 1 
%   |  f()
%   |  |  x changes, scope 2
%   |  |  f()
%B  |  |  |  x changes, scope 3 
%C  |  ... , scope 1
%\end{lstlisting}
%\end{verbatim}
\caption{A recursive call trace illustrating the scope
  id. The lines in the bottom half of the diagram simulate a trace of the change events for the variable \texttt{x} as the function \texttt{f()} calls itself. Each call creates a scope; eventually when the variable \texttt{stop} is changed by an external process we return from the recursion. The lines marked A, B, and C are discussed in the text.}
\label{fig:recursive}
\end{figure}

Having the scope chain and the source code, we can map every scope to % perhaps needs more explanation, how?
a code block, enclosed the executing code. In
JavaScript, a code block can be identified by the file url and  
the block's first instruction program counter. Given this information, the debugger is
able to recognize the code block in loaded scripts or once it is loaded.
Similar to \textit{lastChange} on object property, we define two basic operations: 
\texttt{scopeId()} -- which returns an integer given a scope -- and \texttt{setVariableChangeHook()} -- which calls its first parameter, a callback function, when a 
variable in its second parameter, a code block, matches the name given in the third parameter.

@@

\begin{figure}[htp]
\centering 
\includegraphics[width=.40\textwidth]{7-foo-changes2.pdf}
\caption{A hypothetical list of variable \texttt{foo} change events. Each column of the list indicates a change event; for each
change event the scope id returns by \texttt{scopeId()} is recorded along with call stack and program state information. 
   The last column having the scope id \texttt{foo} at the reproduction point indicates 
  the last change.}
\label{fig:foo-changes2}
\end{figure}

\subsection{{\large\bf\textit{lastChange}} on {\large\bf\textit{lastChange}}}
The \textit{lastChange} algorithm records changes by id (either object or scope id), then reads out the last change when we
arrive at the reproduction point and discover the id of the requested value.
When we perform \textit{lastChange} based on a previous \textit{lastChange}, the query algorithm 
must retain additional information. Consider the following example:

@@

%---------------------------------------------------------------------------------------------------
\section{JavaScript implementation}
To verify the \textit{lastChange} algorithm we
implemented\footnote[1]{http://code.google.com/p/querypoint-debugging} it in an extension to the Firebug
JavaScript debugger\footnote[2]{http://getfirebug.com}. %\cite{Firebug}. %\cite{firebug-version-1.6}. 
Firebug itself is an extension of the Firefox browser. %\cite{Firefox}. %\cite{firefox-version-3.6}
The Firefox JavaScript engine provides a JavaScript debugging interface \cite{JSD} and 
\textit{Querypoint} is developed over this interface. Our prototype implements the four primitive operations in JavaScript 
using techniques which are cumbersome and comparatively slow to execute. However the JavaScript prototype is
easy to explore, change, and share with others for feedback.
  A professionally useful debugger would implement these primitive operations within the JavaScript engine.

\begin{figure}[htp]
\centering 
\includegraphics[width=.48\textwidth]{8-lastchange-lastchange.pdf}
\caption{The list of change events stored for locating point B, the
  \textit{lastChange} of \texttt{bar.x} at the reproduction point, and
  point C, the \textit{lastChange} of \texttt{baz.y} at point B.}
\label{fig:lastchange-lastchange}
\end{figure}

\subsection{{\normalsize\bf\texttt{objectId()}} operation}
\texttt{objectId(obj)} first checks the argument \texttt{obj} for a property \texttt{\_objectId}.
It returns the value if this property is already defined,
otherwise it generates a new id and sets this property. The value of the id is simply an integer incremented 
for each new \texttt{\_objectId} needed. 
%\footnote[3]{This use of a regular object property  might change
%the program behaviour. For example it will appear as an extra property in 
%\texttt{for(property in object)} loops. 
%The next version of Firefox (https://developer.mozilla.org/en/JavaScript/Reference\\/Global\_Objects/Object/defineProperty) provides \texttt{defineProperty}, 
%allowing a property to be added which is invisible to enumeration.} Note that \texttt{\_objectId} is
%not set for all objects but only those objects that need an id.

@@

For our purposes, the \texttt{watch()} function only covers the case
of global object properties. At the beginning of execution, no object
excepting the global object and its predefined properties is available.  For 
our \textit{lastChange} prototype, we created a version of
\texttt{setPropertyChangeHook()}.  The basic strategy is to get a reference to the object just
after its creation, then use \texttt{watch()} function to monitor
property changes in the object. Setting a flag\footnote[4]{DISABLE\_OBJECT\_TRACE defined in jsdIDebuggerService.idl} into the Firefox 
JavaScript engine, we can get file URL and line number for each object
creation (e.g., myFile.js, line 24). We set a breakpoint on this line
and parse the source code to determine which object was created.

@@

\subsection{{\normalsize\bf\texttt{scopeId()}} operation}
The prototype sets a breakpoint at the beginning of all code blocks needing
a scope id. The scope id is kept as a variable with name 
\texttt{\_scopeId} in the scope. Whenever the hook is hit,  meaning a new scope
is created, \texttt{\_scopeId} is set by calling JavaScript's dynamic compilation 
function \texttt{eval()}. For example, executing \texttt{eval("var \_scopeId = 10")} creates a
variable with name \texttt{\_scopeId} and value \texttt{10} in the
scope of the \texttt{eval()} call, which is our interesting scope. \texttt{scopeId()}
operation returns the value of \texttt{\_scopeId} in the scope.

\subsection{{\normalsize\bf\texttt{setVariableChangeHook()}} operation}
Variables defined in local, closure, and catch scopes are only changed in
their scope; that includes the defining scope and scopes nested whithin them.
For example in Figure~\ref{fig:js-closure}, variable \texttt{myVar} in line 11 can only be
changed in lines 10 to 23. Therefore, after locating the function in
which the variable is defined, it is enough to parse the code inside
the function block and set a hook on all lines where the variable is
assigned a new value. 
We also set a hook at the first line of the function which is
corresponding to the line where the variable is defined. In Figure
~\ref{fig:js-closure}, if the execution is paused at line 17 and the
last change of \texttt{myVar} is queried, two hooks on lines 16 and 17
will be set, but if the execution is paused at line 20 , four hooks on
lines 11, 12, 14, 20 will be set. These two cases are different: \texttt{myVar} in 
\texttt{firstChild} is a local variable
but in \texttt{secondChild} it is a closure variable.

@@

% Cut to reduce size
%There are cases where event indexes fail, but a developer can use knowledge gained from \textit{lastChange}
%to formulate a conditional breakpoint.
%Consider the example in Figure~\ref{fig:counter-example}. It shows a function which processes an
%array. The array contains numbers except one item which is
%\texttt{undefined} and it causes a bug in line 30. There is a call to
%\texttt{randomPermutation} function in line 11 which randomly
%permutes the array item. So at every execution the
%\texttt{undefined} item will be in a new place. Calling
%\textit{lastChange} on \texttt{x} at the place the bug happens, gives a
%point which shows line 14. Although this point exists at every
%re-execution but it can not be identified by an event index. A conditional breakpoint on 
%the value of the item would succeed and the developer can reason from the 
%textit{lastChange} results to create the condition.

%\begin{figure}[htp]
%\lstset{basicstyle=\scriptsize}
%\begin{lstlisting}[language=myLang, framerule=0pt]
%10 function randomProcess(array){
%11   randomPermutation(array);
%12   var x, y;
%13   for (var i=0 ; i<array.length ; i++){
%14      x = array[i]
%...
%30      y = x+1;
%31   }
%32 }
%\end{lstlisting}
%\caption{A counter-example for transforming \textit{lastChange} to a conditional breakpoint.}
%\label{fig:counter-example}
%\end{figure}

\section{User Study}
We supplied four experienced Javascript developers with our prototype in an 
extended Firebug debugger\footnote[5]{http://ltiwww.epfl.ch/\texttildelow mirghase/lastchange-userstudy}. Following a tutorial and a practice case, we observed as they 
applied both conventional breakpoint and \textit{lastChange} on two small programs we 
provided (Our prototype did not have a user interface to support Sec.~\ref{sec:pausing}).
All four developers successfully applied \textit{lastChange} to the test programs 
and understood how it could help debugging.

We designed the example programs to require two steps of \textit{lastChange}; 
all four users took more steps (9, 13, 12, and more than 30) to find the defect location with breakpoints.  
To find locations for breakpoints two users scrolled through the source, 
another searched the text, a third set a lot of breakpoints to get an understand of the control flow. 
Based on our own experience we expect these strategies represent the kinds of approaches 
developers have available. These operations are time consuming and tedious. In contrast, all four users 
found the defect location with just two \textit{lastChange} operations.
% Based on our measurements,
%this makes \textit{lastChange} about three times faster than breakpoints for the navigation from 
%the reproduction point to the defect
We recognize that these programs were designed to highlight lastChange and many kinds of debugging 
issues have been hidden by the design of our tests. Nevertheless our results show that, when a defect 
relates to incorrect values and a developer recognizes this, then the operational 
mechanics of lastChange lead to the defect much more quickly than breakpoints.

@@

%The key ingredients in our argument are: 
%\begin{enumerate}
%   \item developers need an operation like \textit{lastChange}, 
%   \item developers can learn to use \textit{lastChange}, 
%   \item practical implementations are feasible with modest
%invested development time,
%   \item in most cases \textit{lastChange} will be much faster than current alternatives,
%   \item the worst cases are not more common or more painful than
%alternatives.
%	 \item the \textit{lastChange} approach can be generalized to other queries.
%\end{enumerate}

@@

%We have shown how \textit{lastChange} identifies the prior point in
%program execution where a program state value changes. Is it something
%the developer needs to do?

Since ultimately programs are just
transformation of state values, debugging is ultimately backtracking
to find defects in program state change\cite{Weiser,Zeller}.  When a developer halts a
program on a breakpoint they compare the call stack and program state to their model of the program.
If any value seems to be incorrect,  they need to figure out what operation
causes the incorrect value. Here \textit{lastChange} takes over and addresses a key
part of the debugging process.
%have three kinds of information: a model
%of the program in their mind, a call stack showing what part of the
%program is halted, and the state at that point (a combination of the
%debugger's view of the program internal state and the output of the
%program up to this point of execution). If the model does not match
%the call stack or if no value in the viewable program state aligns
%with the developers model for this point in execution, the developer
%will seek another view by changing the breakpoint or the input
%data. If a value appears incorrect, they may have a flash of insight
%and know the defect. Otherwise they need to figure out what operation
%causes the incorrect value. Here \textit{lastChange} takes over and addresses a key
%part of the debugging process.

@@

%If \texttt{lastChange} could be useful, can developers figure out how
%to use it? 
While our user study was small, our prototype demonstrates that
\textit{lastChange} can be easily activated by operations on the
graphical representation of erroneous data in a debugger and the 
results can be interpreted by users. 
While \textit{lastChange} is not a breakpoint, 
all of the assumptions developers already have for breakpoints hold
for \textit{lastChange}. In particular the re-execution is just the
same operation developers use to debug with breakpoints.

%We designed the
%user interface for the results from \textit{lastChange} to resemble
%the results from a breakpoint, with the source code of the change
%point highlighted and the state of the program presented the way state
%is presented from a breakpoint. Based on this user interface, we
%believe developers can start to use \textit{lastChange} with minimal
%training. Our users may not be representative, but the key point here
%is that the learning curve is small compared with the challenge of learning 
%breakpoint debugging in the first place.

%In our own experience, \textit{lastChange} dramatically reduces the work in
%setting and removing breakpoints, so work to cause re-execution  become noticable.
%Therefore automatic mechanisms
%for re-execution will be corresponding more valuable for debugging
%with \textit{lastChange} (Our \textit{Querypoint} prototype  
%implements a simple form of automatic 
%re-execution for both breakpoint and \textit{lastChange}).

@@

We have described our prototype all-JavaScript implementation in Sec. 4. It is adequate for 
exploring the ideas and may even be usable in production. However significant 
improvments can be made. A fully usable
implementation would require access to the object id at the point of
object creation and \texttt{setPropertyChangeHook()}. Many
object-oriented runtimes provide object identifiers and 
provide access to object creation directly or by bytecode instrumentation\cite{JPDA}. 
%In the particular case of the Firefox Web browser we used for our prototype, the primary barrier
%to practical implementation would likely be integration with the
%increasingly sophisticated just in time compilers. In the case of
%tracking variable changes, \texttt{setVariableChangeHook()} can be implemented
%similar to \texttt{setPropertyChangeHook()} if JavaScript scopes-like regular 
%JavaScript objects-support \texttt{watch()} function.

@@

%When developers try \textit{lastChange}, will they get results fast
%enough to benefit? While we only used our prototype on toy programs,
%\textit{lastChange} seemed as fast as breakpoint re-execution.
%But what can we say about realistic programs?

@@

%We implemented a
%simple and effective mechanism allowing developers to control 
%overhead: we stored data to a depth of two property accesses 
%(e.g. foo.bar.x), then placed a \textit{re-run} button 
%in the user interface at the third level. If developers reached the 
%third level they can see more data at the cost of one more replay.

For comparison consider today's practical alternative: developers
setting breakpoints. For the vast majority of programs, a developer
will take much more time to set one breakpoint than
\textit{lastChange} would add. Moreover, typically the developer may not
guess the point of last change. They must then ponder another
breakpoint and re-execute.

%Unlike the re-execution we use to 
%optimize performance in \textit{lastChange}, 
%developers still will not know if the breakpoint they hit is the point of last change
 %or even if it is related to the change at all.

@@

%Finally we consider the worst cases: what about code that changes
%objects in long running loops? 
Every time through the loop we incur
the call back overhead; if the loop itself has relatively little code
the overhead could be very large; if the loop computation is a
significant fraction of the full program, the slowdown would be
enormous. 
Other techniques also struggle with this case:  breakpoints in highly repeating loops are not feasible and logging
becomes unwieldy. 
Developers face this issue with any debugger today: occasionally a
debugger causes too much overhead to be useful for debugging. 
%Since automatic logging solutions are highly tuned,
%their overhead in this case will likely be much less than
%\textit{lastChange}. On the other hand, \textit{lastChange} integrates
%with an interactive debugger and detecting that we are in a high
%overhead loop is simply a matter of checking our internal counter. %
%In
%such unusual cases we may simply offer the developer the option of
%studying the loop code then omitting it from \textit{lastChange}
%operations. 
%
%The users of \textit{lastChange} may encounter these kinds of cases, but 
%that is the nature of debugging: you need different tools for 
%different kinds of problems.

\subsection{Generalizations}
Our \textit{lastChange} algorithm can be viewed as a particular interface to a general facility. 
The general facility replays execution, queries the runtime at points of interest during execution,
and analyzes the result at the reproduction point. We have selected one kind of query and analysis
that can be easily integrated in existing debuggers and explained to developers, providing automation of 
the problem of finding the point of last change. We believe other
kinds of queries and analysis can be invented and integrated to automate other aspects of debugging.
%Moreover, the query history analysis used to test for non-determinism here may also be the basis for
%more effective user interfaces to debugging sessions. The series of cascaded \textit{lastChange} queries becomes 
%logical path followed by the developer seeking the defect and this path may be useful if presented in effective user interfaces.
